---
title: "QC Results"
format: 
  html:
    embed-resources: true
execute: 
  echo: false
---


```{r, message=FALSE, echo=FALSE, include=FALSE}
library(tidyverse)
library(fiveneedlepine)

# Import Data

data_dir = "data" 

# Extract the path of the first Access database in the data folder folder of the project directory
# Or the database path can be input manually
database_dir <- common::file.find(here::here("data", "UCBN"), "*.accdb", up = 0)[1]

# Load in all data
pine <- fiveneedlepine::loadPine(database_dir)

# Filter for only SIEN data
pine$data$Seedling <- pine$data$Seedling %>% filter(network == 'SIEN')
pine$data$Site <- pine$data$Site %>% filter(network == 'SIEN')
pine$data$Tree <- pine$data$Tree %>% filter(network == 'SIEN')
pine$data$Visit <- pine$data$Visit %>% filter(network == 'SIEN')

# Read only SIEN data to CSVs
dataPath <- here::here("data", "SIEN")
writePine(here::here("data", "SIEN"), dictionary_dir = here::here("data", "SIEN", "dictionary"))
readr::write_csv(pine$data$Visit, file = paste0(dataPath, "/Visit.csv"))
readr::write_csv(pine$data$Seedling, file = paste0(dataPath, "/Seedling.csv"))
readr::write_csv(pine$data$Tree, file = paste0(dataPath, "/Tree.csv"))
readr::write_csv(pine$data$Site, file = paste0(dataPath, "/Site.csv"))

# load in only SIEN Data
#database_dir <- common::file.find(here::here("data", "SIEN"), "*.accdb", up = 0)[1]
fiveneedlepine::loadPine(here::here("data", "SIEN"))
```

# Event QC

##### Check there is Seedling Data for all Events
```{r}
if(nrow(fiveneedlepine:::noSeedlingDataQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::noSeedlingDataQC())
}
```



##### Check there is Tree Data for all Events
```{r}
if(nrow(fiveneedlepine:::noTreeDataQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::noTreeDataQC())
}
```


# Tree QC

##### Check for Duplicate Tree Tags
```{r}
if(nrow(fiveneedlepine:::treeDuplicateTagQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::treeDuplicateTagQC())
}
```


##### Check for Missing Tree Tags
```{r}
if(nrow(fiveneedlepine:::treeMissingTagQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::treeMissingTagQC())
}
```


##### Check that all the Stem Letters are Letters
```{r}
if(nrow(fiveneedlepine:::stemLetterQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::stemLetterQC())
}
```


##### Check if Cause of Death is missing or a non-Valid Cause of Death
```{r}
if(nrow(fiveneedlepine:::treeCauseOfDeathQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::treeCauseOfDeathQC())
}
```


##### Check if Height is Missing or Three Standard Deviations Greater than Median
```{r}
# TODO: a possible way to nicely print the longer graphs
# DT::datatable(fiveneedlepine:::treeHeightQC() %>% select(-eventID), 
#          #extensions = c('FixedColumns',"FixedHeader"),
#          rownames = FALSE,
#           options = list(scrollX = TRUE, 
#                          paging=TRUE,
#                          fixedHeader=TRUE))
```


```{r}
if(nrow(fiveneedlepine:::treeHeightQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::treeHeightQC())
}
```


##### Check Height Difference

Check for Trees that have a height change greater than 33% between visits 

```{r}
if(nrow(fiveneedlepine:::treeHeightDifferenceQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::treeHeightDifferenceQC())
}
```


##### Check if DBH is Missing or Three Standard Deviations Greater than Median
```{r}
if(nrow(fiveneedlepine:::dbhQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::dbhQC())
}
```

##### Check DBH Difference

Check for Trees that have a DBH change greater than 33% between visits 

```{r}
if(nrow(fiveneedlepine:::dbhDifferenceQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::dbhDifferenceQC())
}
```


##### Check for Recently Dead PIAL Trees with a Null Mortality Year
```{r}
if(nrow(fiveneedlepine:::mortalityYearQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::mortalityYearQC())
}
```


##### Check Cone Count

Return list of trees where female cones exist but there is no cone count or where female cones do not exist but cone count is populated

```{r}
if(nrow(fiveneedlepine:::coneCountQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::coneCountQC())
}
```


##### Check Crown Health

Return list of trees who's crown health isn't a domain value or live PIAL trees with missing crown health

```{r}
if(nrow(fiveneedlepine:::crownHealthQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::crownHealthQC())
}
```


##### Check Lower Crown Kill

Return a list of trees whose lower crown kill is more than 100% or alive PIAL trees that have a null lower crown kill but are not null in both upper and middle crown kill percents

```{r}
if(nrow(fiveneedlepine:::crownKillLowerQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::crownKillLowerQC())
}
```


##### Check Middle Crown Kill

Return a list of trees whose middle crown kill is more than 100% or live PIAL trees that have a null middle crown kill but are not null in both upper and lower crown kill percents

```{r}
if(nrow(fiveneedlepine:::crownKillMiddleQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::crownKillMiddleQC())
}
```


##### Check Upper Crown Kill

Return a list of trees whose upper crown kill is more than 100% or live PIAL trees that have a null upper crown kill but are not null in both middle and lower crown kill percents

```{r}
if(nrow(fiveneedlepine:::crownKillUpperQC()) == 0){
  cat('No errors')
} else{
 rmarkdown::paged_table( fiveneedlepine:::crownKillUpperQC())
}
```


##### Check Subplot Numbers are between 1 and 5
```{r}
if(nrow(fiveneedlepine:::treeSubplotQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::treeSubplotQC())
}
```


##### Check if Vitality is Missing or a non-Valid Domain Value
```{r}
if(nrow(fiveneedlepine:::treeVitalityQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::treeVitalityQC())
}
```


##### Check Lower Bole Cankers

Return a list of trees with non-valid responses in the lower bole canker columns

```{r}
if(nrow(fiveneedlepine:::boleCankersILowerQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::boleCankersILowerQC())
}
```


##### Check Middle Bole Cankers

Return a list of trees with non-valid responses in the middle bole canker columns

```{r}
if(nrow(fiveneedlepine:::boleCankersIMiddleQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::boleCankersIMiddleQC())
}
```


##### Check Upper Bole Cankers

Return a list of trees with non-valid responses in the upper bole canker columns

```{r}
if(nrow(fiveneedlepine:::boleCankersIUpperQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::boleCankersIUpperQC())
}
```


##### Check Lower Branch Cankers

Return a list of trees with non-valid responses in the lower branch canker columns

```{r}
if(nrow(fiveneedlepine:::branchCankersILowerQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::branchCankersILowerQC())
}
```


##### Check Middle Branch Cankers

Return a list of trees with non-valid responses in the middle branch canker columns

```{r}
if(nrow(fiveneedlepine:::branchCankersIMiddleQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::branchCankersIMiddleQC())
}
```


##### Check Upper Branch Cankers

Return a list of trees with non-valid responses in the upper branch canker columns

```{r}
if(nrow(fiveneedlepine:::branchCankersIUpperQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::branchCankersIUpperQC())
}
```


##### Check Species

Return list of trees with missing or blank scientific names

```{r}
if(nrow(fiveneedlepine:::treeSpeciesQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::treeSpeciesQC())
}
```


##### Check that Trees aren't listed as Recently Dead Twice
```{r}
if(nrow(fiveneedlepine:::recentlyDeadTreeQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::recentlyDeadTreeQC())
}
```



# Seedling QC


##### Check for Plots that do not have 9 Subplots
```{r}
if(nrow(fiveneedlepine:::numberOfSubplotsQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::numberOfSubplotsQC())
}
```


##### Check Subplot Numbers are between 1 and 9 
```{r}
if(nrow(fiveneedlepine:::subplotQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::subplotQC())
}
```


##### Check for Missing Tags
```{r}
if(nrow(fiveneedlepine:::missingTagQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::missingTagQC())
}
```


##### Check for Duplicated Tags
```{r}
if(nrow(fiveneedlepine:::duplicateSeedlingTagQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::duplicateSeedlingTagQC())
}
```


##### Check that Cause of Death is a Valid Domain Value
```{r}
if(nrow(fiveneedlepine:::causeOfDeathQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::causeOfDeathQC())
}
```


##### Check that Vitality is a Valid Domain Value
```{r}
if(nrow(fiveneedlepine:::vitalityQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::vitalityQC())
}
```


##### Check Species

Return list of trees with missing or blank scientific names

```{r}
if(nrow(fiveneedlepine:::seedlingSpeciesQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::seedlingSpeciesQC())
}
```


##### Check Height Classes

Return a list of seedlings that are alive but the height class does not match domain values

```{r}
if(nrow(fiveneedlepine:::heightClassQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::heightClassQC())
}
```


##### Check that Seedlings aren't listed as Recently Dead Twice
```{r}
if(nrow(fiveneedlepine:::recentlyDeadSeedlingQC()) == 0){
  cat('No errors')
} else{
  rmarkdown::paged_table(fiveneedlepine:::recentlyDeadSeedlingQC())
}
```









